\sys: Controllable Program Regeneration

### 2025-06-15 paper/draft.tex
Fix weasel and passive voice
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

% \sys is a framework for regenerating programs.
% \sys accepts a variety of wanted program properties that should be preserved after regeneration.
% \sys also takes a set of guiding properties that will be given as input to the regeneration component.
% Then, \sys will regenerate the program to satisfy the wanted properties while preserving the guiding properties.
% \sys can be used for source to source transformation, program repair, (original code can be: insecure, slow, proprietary, wrong, bad context, different language, non-idiomatic).
% \sys uses a plugin architecture that allows extending it with program property extractors (IO pairs, CFG, AST, execution traces, execution time, etc.).
% The regeneration component can be extended with different regeneration algorithms (e.g., genetic programming, symbolic execution, LLM, etc.).

@@

\section{Introduction}
\begin{figure}[t]
  \includegraphics[width=\columnwidth]{figs/sys_overview-figure.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are preserved and unwanted properties are avoided.
}
\end{figure}

@@

Recent advances in large language models (LLMs) have opened new possibilities for automating program transformation and synthesis tasks, making them attractive tools for software evolution, albeit with caveats.
This flexibility has positioned them as attractive tools for software evolution.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
However, their outputs are shaped by surface-level patterns in the input~\cite{yang2025evaluatinggeneralizationcapabilitieslarge},
they struggle with negative constraints~\cite{hwang2024thinkpinkelephant,jiang2024llmsdreamelephantswhen},
often overfit to spurious input cues~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode},
and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{roh2025breakthechainreasoningfailuresllms}.
As a result, developers must prompt and steer them through brittle heuristics or manual trial-and-error, leading to workflows that are difficult to audit, integrate, or systematically reason about.
Bridging this gap calls for a declarative interface---one that allows developers to specify explicit semantic constraints and transformation goals, and to reason about their satisfaction in a principled way.

@@

The design of \sys is shaped by three key insights.
First, modern software is highly modular, making it feasible to reason about transformations at the granularity of individual components or functions.
Second, the program regeneration setting provides a unique advantage: the original program serves as a concrete ground-truth, enabling semantic comparisons and verification of the transformed output, something impossible in traditional synthesis tasks.
Third, the underlying synthesis technology---particularly LLMs---is evolving rapidly; a declarative interface allows this progress to be harnessed without disrupting the framework's end user.

@@

\heading{Deployment scenarios and limitations}
\sys can be deployed at various points in the software lifecycle.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating third-party modules that must be adapted to local conventions or hardened against security vulnerabilities.
However, \sys is not designed for whole-program synthesis or large-scale migrations.
Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module---where transformation goals can be explicitly stated and verified in isolation.
To this end, \sys is designed not only as a standalone tool but also as a library component within larger, compound AI systems~\cite{compound-ai-blog}, where it can be invoked as a programmable, constraint-guided regeneration module.

@@

\sys is evaluated across a diverse set of transformation tasks~(\cref{sec:evaluation}), including:
	security hardening, by stripping latent malicious behavior while preserving intended functionality;
	cross-language translation, by regenerating semantically equivalent programs in new languages;
	idiomatic rewriting, by producing safer and more portable variants of legacy code;
	modular decomposition, by restructuring monolithic components into composable parts.
\sys is able to effectively regenerate programs across these tasks, producing high-quality outputs that satisfy user-specified constraints while also discovering transformation strategies that were not immediately obvious to the authors.

@@

\section{Example}
\label{sec:example}
\sys enables users to specify transformation goals declaratively.
These goals are expressed as logical constraints over properties of the
regenerated program, such as input-output equivalence, side-effect elimination,
target language, lines of code, and others.

Given a program and a query over desired properties, \sys selects a minimal set
of properties to extract from the original program, then attempts to synthesize
a new program that satisfies the query.
This process is cast as a
maximum-satisfiability (MaxSAT) problem~\cite{sung2006maximum}, optimizing for the smallest extraction
effort necessary to enable transformation.
If the query is unsatisfiable under
the constraints or transformations, \sys issues an error
message, showing the conflict.

This section goes through several examples applying \sys
to three distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

\heading{A compromised javascript library}
The \texttt{flatmap-stream} library was implicated in a high-profile
supply-chain attack that exfiltrated Bitcoin credentials~\cite{ev:eurosec:2022}.
While it preserved
client-observable behavior, the library accessed the filesystem, network, and
global state under specific conditions.
Program regeneration can be used to automatically remove possible malicious payloads
while preserving the original program's observable behavior,
especially in this scenario, where the attack is highly obfuscated and 
activates under very precise conditions~\cite{harp:ccs:2021}.
\begin{minted}[frame=lines]{js}
module.exports = function (stream, fn) {
 if (process.env.PRODUCTION) {
  const priv = copay.getKeys()?.[0]?.priv;
  if (priv && wallet.balance > 0) {
    http.request({ hostname: 'evil.com' })
        .send({ priv });
  }} return flatten(steam).apply(fn); };
\end{minted}

\begin{wrapfigure}[3]{r}{.5\columnwidth}
\vspace{-10pt}
\begin{minted}{prolog}
equal(io(P), io(P_)).
pure(P_).
\end{minted}
\end{wrapfigure}
A program written in the \sys DSL to express this intent is shown.
This query requires the regenerated program \ttt{P_} to exhibit the same I/O behavior,
but forbids side-effects.
\sys first arrives at a minimal set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{\texttt{funcs(F, P), io(F), sig(F), pure(P)}\}$, where
\texttt{funcs(F, P)} extracts the set of functions defined in the original program
using a language-aware component that either parses the program source or extracts
symbols from the binary,
\texttt{io(F)} extracts the I/O behavior of each function
by providing its source code to an LLM instance and asking it to generate a set of
inputs; with corresponding outputs generated by executing the original program,
and \texttt{sig(F)} extracts each function's signature.
The, \texttt{pure(P)} property means that \sys will confirm after regeneration that the original program is (obviously) pure
using static analysis.

@@

Building on the same source, \sys can be used for cross-language
translation. 
Here, the goal is to regenerate a semantically equivalent version
of \texttt{flatmap-stream} in Haskell, to enable integration with a
Haskell-based pipeline or facilitate formal reasoning.

@@

The following query for \sys uses lines-of-code a coarse metric of
idiomaticity. \sys will regenerate the program, minimizing this metric, while
preserving its I/O behavior.
\begin{wrapfigure}[3]{r}{.5\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
equal(io(P), io(P_)).
#min len(P_).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties $\{\texttt{sig(F)}, \texttt{io(F)}\}$ from the original program, where \texttt{sig(F)} is the
signature of the function (including its name, return type, and arguments), and \texttt{io(F)} is the I/O behavior of the function
on a set of generated inputs.
In this case, where it is instructed to optimize an aspect of the program, \sys will re-prompt
the model to regenerate the program until no improvements are made or there is a negligible difference after two consecutive iterations.
After \xxx seconds, and \xxx iterations, \sys produces the following program:
\begin{minted}[frame=lines]{c}
#include <math.h>
float Q_rsqrt(float number) {
    return 1.0f / sqrtf(number);
}
\end{minted}
The regenerated version is easier to verify, portable across compilers, and
avoids undefined behavior.
A user could have also provided performance requirements using 
the \ttt{time(P)} property to further guide the regeneration process.
In this case, \sys would have run the two functions inside a tight loop and 
would have aborted the regeneration process if the performance of the generated
program was not within an acceptable user-defined threshold.

@@

Having the database initialization being in the same function as 
the database access has architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
\begin{wrapfigure}[5]{r}{.7\columnwidth}
  \vspace{-10pt}
  \begin{minted}{prolog}
db('music.db').
equal(trace(P,sql),trace(P_,sql)).
func(F1). func(F2).
equal(F2(F1, _), P).
\end{minted}
\end{wrapfigure}
\sys, given the query to the right, is instructed 
to generate program that consists of two functions,
whose composition is equivalent to the original program, and feature 
the same I/O and trace properties (a trace here being the sequence of SQL 
queries transmitted to the database).
\sys wraps the node process with monitors that extract the SQL queries 
sent to the database for \xxx generated inputs.
Then, after \xxx seconds and \xxx attempts, it arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1, \_)$, is equivalent to the original \ttt{getAlbumsByArtist} function across these inputs.
A final processing step renames functions $f_1$ and $f_2$ by prompting an LLM instance.
Alternatively, the user could have provided specific signatures and names for each 
functions as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function connectDB() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'}); 
}
function getAlbumsByArtist(db, artist) {
 const rows = await db.all("SELECT album
 FROM songs WHERE artist = ?", artist);
 return rows.map(row => row.album); }
\end{minted}

@@

\heading{Plan construction}
First, each of the available plugins contributes a set of pre and pos-conditions 
to the logic engine.
Essentially, these describe what needs to hold for an analysis to be applicable to the given 
program and what properties will be preserved or transformed after regeneration.
Then, one or more user-provided regeneration goals (termed as the query), are passed to the planning component.
The full set of rules and predicates are used to construct an ASP program,
which is passed into a solver.
From the solver's output, the planner constructs a plan that specifies which
analysies should be applied to the input program and what properties can be
preserved during regeneration in order to fullfill the query.

\heading{Property extraction} \sys then continues by analyzing the input program to
extract the collection of properties.
These may describe the program's interface
(\eg, exported functions), its operational constraints (\eg, disallowing
filesystem access), or its structure (\eg, use of a particular API or programming idiom).
While some properties are derived from generic analyses
applicable across languages~(\eg examples of input-output behavior), others are tailored to the specifics of the input program's language and runtime model.

@@

\heading{Verification and feedback}
The final phase verifies that the
synthesized program satisfies the target properties.
If verification fails---due to a violated constraint or behavioral divergence---\sys uses this
failure as a signal to refine the plan or resample the synthesis step.
This forms a feedback loop: planning, synthesis, and verification interact until a
satisfactory program is produced or the system exhausts its search.

@@

\heading{Property predicates}
Each property is represented as a logical predicate over the program symbol $p$. Common examples include:
\begin{itemize}
  \item \texttt{language}$(p, \texttt{javascript})$: the input program is written in JavaScript.
  \item \texttt{signature}$(p)$: the program exposes a well-formed function signature.
  \item \texttt{pure}$(p)$: the program does not have side-effects (\eg file system access, network calls, \etc).
\end{itemize}

Properties are classified into three categories:
\begin{itemize}
  \item $\mathsf{goal}(X)$: the user requires $X$ to hold in the regenerated program. Such facts are derived from the user query.
  \item $\mathsf{can}(X)$: the system is capable of extracting or inferring property $X$ from the input. These facts are added by plugins and represent the system's capabilities.
  \item $\mathsf{do}(X)$: $X$ is selected for enforcement and will be preserved or regenerated. The logic engine determines which properties to enforce based on the user query and the system's capabilities.
\end{itemize}

\heading{Planning over properties}
The logic engine determines the set of active properties to enforce by solving a constrained optimization problem over $\mathsf{do}(\cdot)$. It satisfies the following constraints:
\begin{align*}
  \mathsf{goal}(X) &\Rightarrow \mathsf{do}(X) &\text{Goals must be satisfied} \\
  \mathsf{do}(X) &\Leftarrow \mathsf{can}(X) &\text{Apply only extractable properties}
\end{align*}
\sys also supports minimization, maximization and elimination of properties.
These are supported through the $\mathsf{goal\_min}$, $\mathsf{goal\_max}$, and $\mathsf{goal\_not}$ and the corresponding $\mathsf{do\_min}$, $\mathsf{do\_max}$, and $\mathsf{do\_not}$ predicates.
Without loss of generality, the text encompasses the four cases under $\mathsf{goal}$ and $\mathsf{do}$, explicitly mentioning the rest when there is an important semantic difference.

Additionally, the system maximizes the number of extracted properties:
\[
\text{maximize} \left|\left\{ X \mid \mathsf{do}(X) \right\}\right|
\]
Essentially, the planner tries to give the synthesizer as much information
as possible, as long as none of the regeneration goals are violated.

\heading{Plugin-defined rules}
Plugins contribute domain-specific inference rules and constraints. For example, a plugin for JavaScript-to-Haskell translation may define:
\begin{align*}
\mathsf{can}(\texttt{translate}_{\texttt{js} \rightarrow \texttt{hs}}(p)) & \\
\mathsf{do}(\texttt{translate}_{\texttt{js} \rightarrow \texttt{hs}}(p)) &\Rightarrow \mathsf{do}(\texttt{language}(p, \texttt{haskell})) \\
\mathsf{do}(\texttt{translate}_{\texttt{js} \rightarrow \texttt{hs}}(p)) &\Rightarrow \mathsf{language}(p, \texttt{javascript})
\end{align*}

Another plugin might contribute $\mathsf{can}(\texttt{signature}(p))$, indicating that the function signature can be extracted.

@@

goal_min(len(p)). % User query
\end{minted}
  \caption{\textbf{Example \sys logic program.}
  This is a lightly simplified version of the logic program that produces the plan for the idiomatization task~(\cref{sec:example}). 
  The program contains facts coming from (1) the input program (it is written in C),
  (2) three plugins (\ttt{len}, \ttt{signature}, and \ttt{js2hs}),
  (3) axioms from the logic engine (\eg a program can have one language, an analysis will be applied only if possible, all goals must be satisfied), and
  (4) the user query (the user wants to minimize the length of the program).
  }
  \label{lst:logic-example}
\end{listing}

\heading{Extensibility}
\sys's property extraction is decentralized: plugins contribute $\mathsf{can}(\cdot)$ facts, and pre- and post-conditions, while the core planner resolves which properties to enforce based on goals, capabilities, and axioms.
This modularity allows language-specific analyses to coexist with general-purpose reasoning in a unified planning layer.

\section{Synthesis Plan}

The synthesis process begins with a set of enforced properties $\{ X \mid \mathsf{do}(X) \}$ selected by the planner (\cref{sec:dsl}). These properties reflect both user goals (e.g., removing side effects, translating to a different language) and structural characteristics of the input program that should be preserved. The planner constructs a transformation plan that specifies, for each relevant unit of code (typically a function), what constraints must hold in the regenerated version.

For instance, a plan for regenerating a specific function $f$ might include:
\(\mathsf{do}(\texttt{language}(f, \texttt{haskell}))\) the function must be written in Haskell, 
\(\mathsf{do}(\texttt{pure}(f))\) the function must be side-effect-free, and 
\(\mathsf{do}(\texttt{signature}(f))\) the function signature must be preserved.
% These constraints are scoped to the transformation unit and inform the synthesis phase directly.

\heading{LLM-guided regeneration}
The synthesis task is delegated to a language model guided by a structured prompt. The prompt is constructed by composing fragments contributed by plugins that are part of the regeneration plan, each of which can contribute natural language instructions relevant to its domain. For example:
\begin{itemize}
  \item A \ttt{signature} plugin may contribute: \emph{\say{The output should implement a function \texttt{sqrt(n: float): float}.}}
  \item A \ttt{length} plugin combined with a minimization goal may prepend: \emph{\say{The output should be as short as possible while preserving the original functionality.}}
  \item A \ttt{javascript2haskell} plugin may prepend: \emph{\say{The output function should be written in Haskell.}}
\end{itemize}

These fragments are combined with a normalized text representations of each of 
the extracted properties.

\heading{Constraint-guided generation}
While \sys delegates code synthesis to a large language model, the planner constrains the generation space in two key ways.
First, the prompt is constructed to admit only transformations that are valid under the current plan.
For instance, if the target function must be pure, the prompt will exclude information such as system-call traces that might imply side-effectful behavior.
Second, candidate outputs are verified post hoc against the enforced properties~(\cref{sec:verification}); if verification fails, the result is either rejected or used to revise the current plan.
% This hybrid model balances flexibility with rigor: the LLM provides diversity and fluency, while the planner ensures adherence to property-based constraints.

% In sum, the synthesis plan acts as a structured contract between logical property reasoning and generative synthesis. By operating at a localized level and under clearly scoped constraints, \sys delivers regenerations that are both controlled and scalable---when applied thoughtfully and systematically.

@@

\heading{Failure-driven refinement}
When a candidate fails verification, \sys treats this as a feedback signal. Instead of discarding the regeneration attempt wholesale, the system uses the failure to revise or re-execute the synthesis phase. This may take several forms:
\begin{itemize}
  \item \emph{Retrying synthesis}: A fresh LLM call with the same plan but a different generation path.
  \item \emph{Feedback loop}: \sys will provide the previous regeneration attempt 
    as an additional context to the LLM, together with an explanation of the properties that were violated, prompting it to adapt its output based on prior failures.
    This mode is only activated if the verifier knows that the output will not to taint
    followupp regeneration attempts.
  \item \emph{Fallback handling}: If a function repeatedly fails validation, \sys may leave the original version intact or mark it for manual inspection.
\end{itemize}

@@

\heading{Partial regeneration and composition}
\sys supports regeneration at the granularity of individual functions. If only a subset of functions can be successfully regenerated under the given constraints, the remaining ones are excluded from transformation.
The output program thus consists of a composition of validated regenerated units and untouched originals.
This model supports application in real-world codebases, where partial improvement (\eg replacing unsafe or deprecated functions) can be meaningful even without complete coverage.

>>>>

@@

\section{Introduction}
\begin{figure}[t]
  \includegraphics[width=\columnwidth]{figs/sys_overview-figure.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
\end{figure}

@@

Recent advances in large language models (LLMs) have opened new possibilities for automating program transformation and synthesis tasks, making them attractive tools for software evolution, albeit with caveats.
This flexibility has positioned them as attractive tools for software evolution.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
However, surface-level patterns in the input shape their outputs~\cite{yang2025evaluatinggeneralizationcapabilitieslarge},
they struggle with negative constraints~\cite{hwang2024thinkpinkelephant,jiang2024llmsdreamelephantswhen},
often overfit to spurious input cues~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode},
and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{roh2025breakthechainreasoningfailuresllms}.
As a result, developers must prompt and steer them through brittle heuristics or manual trial-and-error, leading to workflows that are difficult to audit, integrate, or systematically reason about.
Bridging this gap calls for a declarative interface---one that allows developers to specify explicit semantic constraints and transformation goals, and to reason about their satisfaction in a principled way.

@@

Three key insights shape the design of \sys.
First, modern software is highly modular, making it feasible to reason about transformations at the granularity of individual components or functions.
Second, the program regeneration setting provides a unique advantage: the original program serves as a concrete ground-truth, enabling semantic comparisons and verification of the transformed output, something impossible in traditional synthesis tasks.
Third, the underlying synthesis technology---particularly LLMs---is evolving rapidly; a declarative interface that harnesses this progress without disrupting the framework's end user.

@@

\heading{Deployment scenarios and limitations}
\sys supports a range of deployment scenarios.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating third-party modules that must be adapted to local conventions or hardened against security vulnerabilities.
However, \sys is not designed for whole-program synthesis or large-scale migrations.
Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module---where transformation goals can be explicitly stated and verified in isolation.
To this end, \sys functions not only as a standalone tool but also as a
library, where other components can invoke it as a programmable,
constraint-guided regeneration module.

@@

We evaluate \sys across a diverse set of transformation tasks~(\cref{sec:evaluation}), including:
	security hardening, by stripping latent malicious behavior while preserving intended functionality;
	cross-language translation, by regenerating semantically equivalent programs in new languages;
	idiomatic rewriting, by producing safer and more portable variants of legacy code;
	modular decomposition, by restructuring monolithic components into composable parts.
\sys is able to effectively regenerate programs across these tasks, producing high-quality outputs that satisfy user-specified constraints while also discovering transformation strategies that were not immediately obvious to the authors.

@@

\section{Example}
\label{sec:example}
\sys enables users to specify transformation goals declaratively.
Logical constraints over properties of the
regenerated program express these goals. Examples are input-output equivalence, side-effect elimination,
target language, lines of code, and others.

Given a program and a query over desired properties, \sys selects a minimal set
of properties to extract from the original program, then attempts to synthesize
a new program that satisfies the query.
\sys casts this process as a
maximum-satisfiability (MaxSAT) problem~\cite{sung2006maximum}, optimizing for the smallest extraction
effort necessary to enable transformation.
If the query is unsatisfiable under
the constraints or transformations, \sys issues an error
message, showing the conflict.

This section shows the application of \sys
to three distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

\heading{A compromised javascript library}
The \texttt{flatmap-stream} library was implicated in a high-profile
supply-chain attack that exfiltrated Bitcoin credentials~\cite{ev:eurosec:2022}.
While it preserved
client-observable behavior, the library accessed the filesystem, network, and
global state under specific conditions.
Program regeneration can automatically remove possible malicious payloads
while preserving the original program's observable behavior,
especially in this scenario, where the attack is highly obfuscated and 
activates under precise conditions~\cite{harp:ccs:2021}.
\begin{minted}[frame=lines]{js}
module.exports = function (stream, fn) {
 if (process.env.PRODUCTION) {
  const priv = copay.getKeys()?.[0]?.priv;
  if (priv && wallet.balance > 0) {
    http.request({ hostname: 'evil.com' })
        .send({ priv });
  }} return flatten(steam).apply(fn); };
\end{minted}

\begin{wrapfigure}[3]{r}{.5\columnwidth}
\vspace{-10pt}
\begin{minted}{prolog}
equal(io(P), io(P_)).
pure(P_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P_} to exhibit the same I/O behavior,
but forbids side-effects.
\sys first arrives at a minimal set of properties to extract from the
original program, given its knowledge base and the query.
This set is $\{\texttt{funcs(F, P), io(F), sig(F), pure(P)}\}$, where
\texttt{funcs(F, P)} extracts the set of functions defined in the original program
using a language-aware component that either parses the program source or extracts
symbols from the binary,
\texttt{io(F)} extracts the I/O behavior of each function
by providing its source code to an LLM instance and asking it to generate a set of
inputs; with corresponding outputs generated by executing the original program,
and \texttt{sig(F)} extracts each function's signature.
The, \texttt{pure(P)} property means that \sys will confirm after regeneration that the original program is (obviously) pure
using static analysis.

@@

Building on the same source, \sys can perform cross-language
translation. 
Here, the goal is to regenerate a semantically equivalent version
of \texttt{flatmap-stream} in Haskell, to enable integration with a
Haskell-based pipeline or facilitate formal reasoning.

@@

The following query for \sys uses lines-of-code a coarse metric of
idiomaticity. \sys will regenerate the program, minimizing this metric, while
preserving its I/O behavior.
\begin{wrapfigure}[3]{r}{.5\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
equal(io(P), io(P_)).
#min len(P_).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties $\{\texttt{sig(F)}, \texttt{io(F)}\}$ from the original program, where \texttt{sig(F)} is the
signature of the function (including its name, return type, and arguments), and \texttt{io(F)} is the I/O behavior of the function
on a set of generated inputs.
In this case, where the goal is to optimize an aspect of the program, \sys will re-prompt
the model to regenerate the program until it can make no improvements or there is a negligible difference after two consecutive iterations.
After \xxx seconds, and \xxx iterations, \sys produces the following program:
\begin{minted}[frame=lines]{c}
#include <math.h>
float Q_rsqrt(float number) {
    return 1.0f / sqrtf(number);
}
\end{minted}
The regenerated version is easier to verify, portable across compilers, and
avoids undefined behavior.
A user could have also provided performance requirements using 
the \ttt{time(P)} property to further guide the regeneration process.
In this case, \sys would have run the two functions inside a tight loop and 
would have aborted the regeneration process if the performance of the generated
program was not within an acceptable user-defined threshold.

@@

Having the database initialization being in the same function as 
the database access has architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
\begin{wrapfigure}[5]{r}{.7\columnwidth}
  \vspace{-10pt}
  \begin{minted}{prolog}
db('music.db').
equal(trace(P,sql),trace(P_,sql)).
func(F1). func(F2).
equal(F2(F1, _), P).
\end{minted}
\end{wrapfigure}
\sys, given the query to the right, will 
generate program that consists of two functions,
whose composition is equivalent to the original program, and feature 
the same I/O and trace properties (a trace here being the sequence of SQL 
queries transmitted to the database).
\sys wraps the node process with monitors that extract the SQL queries 
sent to the database for \xxx generated inputs.
Then, after \xxx seconds and \xxx attempts, it arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1, \_)$, is equivalent to the original \ttt{getAlbumsByArtist} function across these inputs.
A final processing step renames functions $f_1$ and $f_2$ by prompting an LLM instance.
Alternatively, the user could have provided specific signatures and names for each 
functions as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function connectDB() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'}); 
}
function getAlbumsByArtist(db, artist) {
 const rows = await db.all("SELECT album
 FROM songs WHERE artist = ?", artist);
 return rows.map(row => row.album); }
\end{minted}

@@

\heading{Plan construction}
First, each of the available plugins contributes a set of pre and pos-conditions 
to the logic engine.
Essentially, these describe what needs to hold for an analysis to be applicable to the given 
program and what properties \sys will preserve or transform after regeneration.
Then, the planning component accepts one or more user-provided regeneration goals (termed as the query).
The full set of rules and predicates are used to construct an ASP program
a solver then accepts.
From the solver's output, the planner constructs a plan that specifies which
analysies \sys plugins should apply to the input program and what properties can be
preserved during regeneration in order to fullfill the query.

\heading{Property extraction} \sys then continues by analyzing the input program to
extract the collection of properties.
These may describe the program's interface
(\eg, exported functions), its operational constraints (\eg, disallowing
filesystem access), or its structure (\eg, use of a particular API or programming idiom).
While some properties are derived from generic analyses
applicable across languages~(\eg examples of input-output behavior), others are specific to the input program's language and runtime model.

@@

\heading{Verification and feedback}
The final phase verifies that the
synthesized program satisfies the target properties.
If verification fails---due to a violated constraint or behavioral divergence---\sys uses this
failure as a signal to refine the plan or resample the synthesis step.
This forms a feedback loop: planning, synthesis, and verification interact until \sys
produces a satisfactory program or the system exhausts its search.

@@

\heading{Property predicates}
Each property is a logical predicate over the program symbol $p$. Common examples include:
\begin{itemize}
  \item \texttt{language}$(p, \texttt{javascript})$: the input program is written in JavaScript.
  \item \texttt{signature}$(p)$: the program exposes a well-formed function signature.
  \item \texttt{pure}$(p)$: the program does not have side-effects (\eg file system access, network calls, \etc).
\end{itemize}

Properties fall into three categories:
\begin{itemize}
  \item $\mathsf{goal}(X)$: the user requires $X$ to hold in the regenerated program. The system derives such facts from the user query.
  \item $\mathsf{can}(X)$: the system is capable of extracting or inferring property $X$ from the input. Plugins add these facts and represent the system's capabilities.
  \item $\mathsf{do}(X)$: $X$ the planner selects $X$ for enforcement and the synthesizer will preserve or regenerate it. 
\end{itemize}

\heading{Planning over properties}
The logic engine determines the set of active properties to enforce by solving a constrained optimization problem over $\mathsf{do}(\cdot)$. It satisfies the following constraints:
\begin{align*}
  \mathsf{goal}(X) &\Rightarrow \mathsf{do}(X) &\text{Satisfy all goals} \\
  \mathsf{do}(X) &\Leftarrow \mathsf{can}(X) &\text{Apply only extractable properties}
\end{align*}
\sys also supports minimization, maximization and elimination of properties
through the $\mathsf{goal\_min}$, $\mathsf{goal\_max}$, and $\mathsf{goal\_not}$ and the corresponding $\mathsf{do\_min}$, $\mathsf{do\_max}$, and $\mathsf{do\_not}$ predicates.
Without loss of generality, the text encompasses the four cases under $\mathsf{goal}$ and $\mathsf{do}$, explicitly mentioning the rest when there is an important semantic difference.

Additionally, the system maximizes the number of extracted properties:
\[
\text{maximize} \left|\left\{ X \mid \mathsf{do}(X) \right\}\right|
\]
Essentially, the planner tries to give the synthesizer as much information
as possible, as long as it doesn't violate any regeneration goals.

\heading{Plugin-defined rules}
Plugins contribute domain-specific inference rules and constraints. For example, a plugin for JavaScript-to-Haskell translation may define:
\begin{align*}
\mathsf{can}(\texttt{translate}_{\texttt{js} \rightarrow \texttt{hs}}(p)) & \\
\mathsf{do}(\texttt{translate}_{\texttt{js} \rightarrow \texttt{hs}}(p)) &\Rightarrow \mathsf{do}(\texttt{language}(p, \texttt{haskell})) \\
\mathsf{do}(\texttt{translate}_{\texttt{js} \rightarrow \texttt{hs}}(p)) &\Rightarrow \mathsf{language}(p, \texttt{javascript})
\end{align*}
Another plugin might contribute $\mathsf{can}(\texttt{signature}(p))$, indicating it can extract the function signature. At this stage, \sys only cares about semantics of each plugin in relation to how it might enable or violate 
other properties in the regenerated program.

@@

goal_min(len(p)). % User query
\end{minted}
  \caption{\textbf{Example \sys logic program.}
  This is a lightly simplified version of the logic program that produces the plan for the idiomatization task~(\cref{sec:example}). 
  The program contains facts coming from (1) the input program (written in C),
  (2) three plugins (\ttt{len}, \ttt{signature}, and \ttt{js2hs}),
  (3) axioms from the logic engine (\eg a program can have one language, an analysis will be applied only if possible, all goals must be satisfied), and
  (4) the user query (the user wants to minimize the length of the program).
  }
  \label{lst:logic-example}
\end{listing}

\heading{Extensibility}
\sys decentralizes property extraction: plugins contribute $\mathsf{can}(\cdot)$ facts, and pre- and post-conditions, while the core planner resolves which properties to enforce based on goals, capabilities, and axioms.
This modularity allows language-specific analyses to coexist with general-purpose reasoning in a unified planning layer.

\section{Synthesis Plan}
\label{sec:synthesis}

The synthesis process begins with a set of enforced properties $\{ X \mid \mathsf{do}(X) \}$
which the planner selected~(\cref{sec:dsl}).
These properties
reflect both user goals (\eg removing side effects, translating to a
different language) and structural characteristics of the input program that
the synthesizer will preserve.
The planner constructs a transformation plan that
specifies, for each relevant unit of code (typically a function), what
constraints must hold in the regenerated version.

For instance, a plan for regenerating a specific function $f$ might include:
\(\mathsf{do}(\texttt{language}(f, \texttt{haskell}))\) the function must be written in Haskell, 
\(\mathsf{do}(\texttt{pure}(f))\) the function must be side-effect-free, and 
\(\mathsf{do}(\texttt{signature}(f))\) the function must keep its signature.

\heading{LLM-guided regeneration}
A structured prompt guides the language model, which performs the synthesis task.
The prompt is a composition of natural-language text fragments plugins contribute. 
A plugin can include domain-specific instructions as further guidance.
For example:
\begin{itemize}
  \item A \ttt{signature} plugin may contribute: \emph{\say{The output should implement a function \texttt{sqrt(n: float): float}.}}
  \item A \ttt{length} plugin combined with a minimization goal may prepend: \emph{\say{The output should be as short as possible while preserving the original functionality.}}
  \item A \ttt{javascript2haskell} plugin may prepend: \emph{\say{The output function should be written in Haskell.}}
\end{itemize}

These fragments, combined with a normalized text representations of each of 
the extracted properties constitute the prompt given to the LLM.

\heading{Constraint-guided generation}
While \sys delegates code synthesis to a large language model, the planner constrains the generation space in two key ways.
First, the prompt instructs the LLM to perform only transformations that are valid under the current plan.
For instance, if the target function must be pure, the prompt will exclude information such as system-call traces that might imply side-effectful behavior.
Second, candidate outputs are verified post hoc against the enforced properties~(\cref{sec:verification}).
% This hybrid model balances flexibility with rigor: the LLM provides diversity and fluency, while the planner ensures adherence to property-based constraints.

@@

\heading{Failure-driven refinement}
When a candidate fails verification, \sys treats this as a feedback signal. Instead of discarding the regeneration attempt wholesale, the system uses the failure to revise or re-execute the synthesis phase.
Some forms are:
\begin{itemize}
  \item \emph{Retrying synthesis}: A fresh LLM call with the same plan but a different generation path.
  \item \emph{Feedback loop}: \sys will provide the previous regeneration attempt 
    as an additional context to the LLM, together with an explanation of the properties that the regeneration violates, prompting it to adapt its output based on prior failures.
    This mode is only activated if the verifier knows that the output will not to taint
    followupp regeneration attempts.
  \item \emph{Fallback handling}: If a function repeatedly fails validation, \sys may leave the original version intact or mark it for manual inspection.
\end{itemize}

@@

\heading{Partial regeneration and composition}
\sys supports regeneration at the granularity of individual functions.
If only a subset of functions can be successfully regenerated under the given constraints, the remaining ones remain unchanged after regeneration.
The output program thus consists of a composition of validated regenerated units and untouched originals.
This model supports application in real-world codebases, where partial improvement (\eg replacing unsafe or deprecated functions) can be meaningful even without complete coverage.
